Refactoring the \texttt{nls()} function in R

### 2022-07-27 RefactoringNLS/RefactoringNLS.tex
Some cleanup of bib file

\texttt{nls()} is the tool in base R, the primary software software
distribution from the Comprehensive R Archive Network
(\url{https://cran.r-project.org}) for estimating nonlinear statistical
models. The function dates to the 1980s and the work related to D. M.
Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}).

@@

This issue can be easily resolved. The ``termination test'' for the
\textbf{program} rather than for ``convergence'' of the underlying
\textbf{algorithm} is the Relative Offset Convergence Criterion (see
Douglas M. Bates and Watts
(\protect\hyperlink{ref-BatesWatts81}{1981})). This projects the
proposed step in the parameter vector on the gradient and estimates how
much the sum of squares loss function should decrease. This estimate is
divided by the current size of the loss function to avoid scale issues.
When we have ``converged'', the estimated decrease is very small.
However, with small residuals, the sum of squares loss function is
(almost) zero and we get the possibility of a zero-divide failure.

@@

The convergence criterion above, the principal termination control of
\texttt{nls()}, leaves out some possibilities that could be useful for
some problems. The package \texttt{nlsr} (John C. Nash and Murdoch
(\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
safeguarded relative offset test (\textbf{roffset}) as well as a
\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

@@

gets over 4000 hits that are spread over some years. This could be
because the Jacobian is poorly approximated (see \textbf{Jacobian
computation} below). However, it is common in nonlinear least squares
computations that the Jacobian is very close to singular for some values
of the model parameters. In such cases, we need to find an alternative
algorithm to the Gauss-Newton iteration of \texttt{nls()}. The most
common work-around is the Levenberg-Marquardt stabilization (see
Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}), Levenberg
(\protect\hyperlink{ref-Levenberg1944}{1944}), John C. Nash
(\protect\hyperlink{ref-jn77ima}{1977})), and versions of it have been
implemented in packages \texttt{minpack.lm} and \texttt{nlsr}. In our
work, we prepared experimental prototypes of \texttt{nls} that
incorporate stabilization, but integration with all the features of
\texttt{nls()} has proved difficult.

@@

Because the Gauss-Newton algorithm is unreliable from many sets of
starting of parameters, selfStart models are a part of the
\texttt{nls()} infrastructure rather than an accessory. However,
creating such functions is a lot of work, and their documentation (file
\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
believe that the focus should be placed on getting good initial
parameters, that is \texttt{getInitial()} function, though avoiding the
current calls back to \texttt{nls()}. Interactive tools, such as
``visual fitting'' (John C. Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering.

@@

\begin{verbatim}
## Using approximation  jacentral   is.character(ctrl$japprox)= TRUE 
## <<lamda: 1e-04  SS= 82.379 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
## <<lamda: 4e-05  SS= 27.961 ( 0.87591 ) at  Asym = 23.072  xmid = 24.605  scal = 33.424 f/j  2 / 1
## <<lamda: 0.0016  SS= 27.179 ( 0.10632 ) at  Asym = 27.088  xmid = 25.624  scal = 24.768 f/j  5 / 2
## <<lamda: 0.0064  SS= 26.326 ( 0.11129 ) at  Asym = 30.015  xmid = 24.489  scal = 21.971 f/j  7 / 3
## <<lamda: 0.00256  SS= 26.011 ( 0.18194 ) at  Asym = 36.922  xmid = 21.283  scal = 12.464 f/j  8 / 4
## <<lamda: 0.01024  SS= 21.446 ( 0.27496 ) at  Asym = 39.825  xmid = 16.241  scal = 10.675 f/j  10 / 5
## <<lamda: 0.04096  SS= 19.865 ( 0.18158 ) at  Asym = 40.684  xmid = 14.706  scal = 8.9148 f/j  12 / 6
## <<lamda: 0.016384  SS= 11.81 ( 0.32805 ) at  Asym = 42.981  xmid = 9.344  scal = 4.7286 f/j  13 / 7
## <<lamda: 0.065536  SS= 3.8548 ( 0.46972 ) at  Asym = 43.701  xmid = 6.3095  scal = 2.7564 f/j  15 / 8
## <<lamda: 0.026214  SS= 2.8386 ( 0.45514 ) at  Asym = 44.458  xmid = 5.503  scal = 2.0895 f/j  16 / 9
## <<lamda: 0.010486  SS= 2.4496 ( 0.2365 ) at  Asym = 46.866  xmid = 5.9479  scal = 2.3 f/j  17 / 10
## <<lamda: 0.0041943  SS= 2.0123 ( 0.21463 ) at  Asym = 51.378  xmid = 6.3516  scal = 2.3762 f/j  18 / 11
## <<lamda: 0.0016777  SS= 1.4462 ( 0.2703 ) at  Asym = 59.51  xmid = 7.03  scal = 2.5092 f/j  19 / 12
## <<lamda: 0.00067109  SS= 0.91272 ( 0.306 ) at  Asym = 71.787  xmid = 7.8952  scal = 2.6532 f/j  20 / 13
## <<lamda: 0.00026844  SS= 0.51816 ( 0.29993 ) at  Asym = 88.001  xmid = 8.8228  scal = 2.7857 f/j  21 / 14
## <<lamda: 0.00010737  SS= 0.26988 ( 0.25879 ) at  Asym = 107.49  xmid = 9.727  scal = 2.901 f/j  22 / 15
## <<lamda: 4.295e-05  SS= 0.13173 ( 0.20437 ) at  Asym = 129.22  xmid = 10.558  scal = 2.9978 f/j  23 / 16
## <<lamda: 1.718e-05  SS= 0.062773 ( 0.16799 ) at  Asym = 151.62  xmid = 11.279  scal = 3.0748 f/j  24 / 17
## <<lamda: 6.8719e-06  SS= 0.032464 ( 0.12324 ) at  Asym = 172.29  xmid = 11.849  scal = 3.1314 f/j  25 / 18
## <<lamda: 2.7488e-06  SS= 0.021923 ( 0.076781 ) at  Asym = 188.07  xmid = 12.235  scal = 3.1673 f/j  26 / 19
## <<lamda: 1.0995e-06  SS= 0.01977 ( 0.035429 ) at  Asym = 196.72  xmid = 12.43  scal = 3.1847 f/j  27 / 20
## <<lamda: 4.398e-07  SS= 0.019619 ( 0.0092109 ) at  Asym = 199.45  xmid = 12.49  scal = 3.19 f/j  28 / 21
## <<lamda: 1.7592e-07  SS= 0.019617 ( 0.00087595 ) at  Asym = 199.84  xmid = 12.499  scal = 3.1908 f/j  29 / 22
## <<lamda: 7.0369e-08  SS= 0.019617 ( 4.7993e-05 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  30 / 23
## <<lamda: 2.8147e-08  SS= 0.019617 ( 8.24e-07 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  31 / 24
\end{verbatim}

@@

One obvious case is that nonlinear modeling tools are a superset of
linear ones. Yet the explicit model
\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the
linear modeling function \texttt{lm()}, which requires this model to be
specified as \texttt{y\ \textasciitilde{}\ x}.

Within \texttt{nls()}, consider the following FOUR different calling
sequences for the same problem, though an intuitive choice, labelled
fm2a, will not work. In this failed attempt, putting the \texttt{Asym}
parameter in the model causes the \texttt{plinear} algorithm to try to
add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing, we do not
foresee a resolution for this issue. In the example, we have NOT
evaluated the commands to save space.

@@

The \texttt{nls()} code is structured in a way that inhibits both
maintenance and improvement. In particular, the iterative setup is such
that introduction of Marquardt stabilization is not easily available. To
obtain performance a lot of the code was in C with consequent calls and
returns that complicate the code. Over time, R has become much more
efficient on modern computers, and the need to use compiled C and
Fortran is less critical. Moreover, the burden for maintenance could be
reduced by moving code entirely to R.

While R and its packages are generally very well documented for usage,
that documentation rarely extends as far as it might in directions
helpful for code maintenance. The paucity of such documentation is
exacerbated by the mixed R/C/Fortran code base, where seemingly trivial
differences in things like representation of numbers or base index for
arrays lead to traps for unwary programmers.

For \texttt{nls()} the concern is demonstrated by a short email query
from John Nash to Doug Bates and his reply. This is NOT a criticism of
Prof.~Bates' (or any other) work, but a reflection on how difficult it
is to develop code in this subject area and to keep it maintainable. We
have experienced similar loss of understanding for some of our own
codes.

\begin{verbatim}
    ...
    https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R
    ... has the test problem and the expanded code. Around line 367 is where we are
    scratching our heads. The function code (from nlsModel()) is in the commented 
    lines below the call. This is

          # > setPars
          # function(newPars) {
          #   setPars(newPars)
          #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x}
          #   dev   <<- sum(resid^2) # envir = thisEnv
          #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr)
          #   QR <<- qr(.swts * gr) # envir = thisEnv
          #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix
          # }

    I'm anticipating that we will be able to set up a (possibly inefficient) code
    with documentation that will be easier to follow and test, then gradually figure
    out how to make it more efficient.

    The equivalent from minpack.lm is

    setPars = function(newPars) {
                setPars(newPars)
                assign("resid", .swts * (lhs - assign("rhs", getRHS(),
                    envir = thisEnv)), envir = thisEnv)
                assign("dev", sum(resid^2), envir = thisEnv)
                assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv)
                return(QR$rank < min(dim(QR$qr)))
            }

    In both there is the recursive call, which must have a purpose I don't understand.
\end{verbatim}

The reply was

\begin{verbatim}
    I'm afraid that I don't know the purpose of the recursive call either.  
    I know that I wrote the code to use a closure for the response, covariates, etc., 
    but I don't recall anything like a recursive call being necessary.
    ...
\end{verbatim}

@@

@@

\begin{itemize}
\item
  to provide a packaged version of \texttt{nls()} (call it
  \texttt{nlsalt}) coded entirely in R that matches the features in base
  R or what is packaged in \texttt{nlspkg} as described in John C. Nash
  and Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022}).
\item
  to streamline the overall \texttt{nls()} infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to
  maintenance of both the existing \texttt{nls()} methods and features
  as well as new features we or others would like to add.
\item
  to explain what we do, either in comments or separate maintainer
  documentation. Since we are complaining about the lack of explanatory
  material for the current code, we feel it is incumbent on us to
  provide such material for our own work, and if possible for the
  existing code.
\end{itemize}

@@

\begin{itemize}
\item
  A BibTex bibliography for use with all documents in this project, but
  which has wider application to nonlinear least squares projects in
  general
  (\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib})
\item
  \texttt{MachID.R} offers a suggested concise summary function to
  identify a particular computational system used for tests. A
  discussion of how it was built and the resources needed across
  platforms is given in at
  \url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}
\item
  John C. Nash and Bhattacharjee
  (\protect\hyperlink{ref-PkgFromRbase22}{2022}) is an explanation of
  the construction of the \texttt{nlspkg} from the \texttt{nls()} code
  in R-base.
\item
  As the 2021 Summer of Code period was ending, one of us (JN) was
  invited to prepare a review of optimization in R. Ideas from the
  present work have been instrumental in the creation of John C. Nash
  (\protect\hyperlink{ref-NashWires22}{2022}).
\end{itemize}

@@

From a design point of view, a key difference in approach between
\texttt{nls()} and \texttt{nlsr::nlxb()} is that \texttt{nls()} builds a
large infrastructure, especially the \texttt{nlsModel()} function, from
which the Gauss-Newton iteration can be executed and other statistical
information such as profiles can be computed, while \texttt{nlxb()}
returns quite limited information, and computes what is needed on an
``as and when'' basis. This follows a path that one of us (JN)
established almost 50 years ago with the software that became John C.
Nash (\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver,
and post-solution analysis phases of computations.

@@

\leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
Bates, Douglas M., and Donald G. Watts. 1981. {``A Relative Offset
Orthogonality Convergence Criterion for Nonlinear Least Squares.''}
\emph{Technometrics} 23 (2): 179--83.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-nlsr-manual}{}}%
Nash, John C, and Duncan Murdoch. 2019. \emph{Nlsr: Functions for
Nonlinear Least Squares Solutions}.

>>>>

\texttt{nls()} is the tool in base R, the primary software software
distribution from the Comprehensive R Archive Network
(\url{https://cran.r-project.org}) for estimating nonlinear statistical
models. The function dates to the 1980s and the work related to Bates
and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}).

@@

This issue can be easily resolved. The ``termination test'' for the
\textbf{program} rather than for ``convergence'' of the underlying
\textbf{algorithm} is the Relative Offset Convergence Criterion (see
Bates, Douglas M. and Watts, Donald G.
(\protect\hyperlink{ref-BatesWatts81}{1981})). This projects the
proposed step in the parameter vector on the gradient and estimates how
much the sum of squares loss function should decrease. This estimate is
divided by the current size of the loss function to avoid scale issues.
When we have ``converged'', the estimated decrease is very small.
However, with small residuals, the sum of squares loss function is
(almost) zero and we get the possibility of a zero-divide failure.

@@

The convergence criterion above, the principal termination control of
\texttt{nls()}, leaves out some possibilities that could be useful for
some problems. The package \texttt{nlsr} (John C Nash and Duncan Murdoch
(\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
safeguarded relative offset test (\textbf{roffset}) as well as a
\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

@@

gets over 4000 hits that are spread over some years. This could be
because the Jacobian is poorly approximated (see \textbf{Jacobian
computation} below). However, it is common in nonlinear least squares
computations that the Jacobian is very close to singular for some values
of the model parameters. In such cases, we need to find an alternative
algorithm to the Gauss-Newton iteration of \texttt{nls()}. The most
common work-around is the Levenberg-Marquardt stabilization (see
Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}), Levenberg
(\protect\hyperlink{ref-Levenberg1944}{1944}), Nash
(\protect\hyperlink{ref-jn77ima}{1977})), and versions of it have been
implemented in packages \texttt{minpack.lm} and \texttt{nlsr}. In our
work, we prepared experimental prototypes of \texttt{nls} that
incorporate stabilization, but integration with all the features of
\texttt{nls()} has proved difficult.

@@

Because the Gauss-Newton algorithm is unreliable from many sets of
starting of parameters, selfStart models are a part of the
\texttt{nls()} infrastructure rather than an accessory. However,
creating such functions is a lot of work, and their documentation (file
\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
believe that the focus should be placed on getting good initial
parameters, that is \texttt{getInitial()} function, though avoiding the
current calls back to \texttt{nls()}. Interactive tools, such as
``visual fitting'' (Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering.

@@

\begin{verbatim}
## Using approximation  jacentral  
##       is.character(ctrl$japprox)= TRUE 
## <<lamda: 1e-04  SS= 82.379 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
## <<lamda: 4e-05  SS= 27.961 ( 0.87591 ) at  Asym = 23.072  xmid = 24.605  scal = 33.424 f/j  2 / 1
## <<lamda: 0.0016  SS= 27.179 ( 0.10632 ) at  Asym = 27.088  xmid = 25.624  scal = 24.768 f/j  5 / 2
## <<lamda: 0.0064  SS= 26.326 ( 0.11129 ) at  Asym = 30.015  xmid = 24.489  scal = 21.971 f/j  7 / 3
## <<lamda: 0.00256  SS= 26.011 ( 0.18194 ) at  Asym = 36.922  xmid = 21.283  scal = 12.464 f/j  8 / 4
## <<lamda: 0.01024  SS= 21.446 ( 0.27496 ) at  Asym = 39.825  xmid = 16.241  scal = 10.675 f/j  10 / 5
## <<lamda: 0.04096  SS= 19.865 ( 0.18158 ) at  Asym = 40.684  xmid = 14.706  scal = 8.9148 f/j  12 / 6
## <<lamda: 0.016384  SS= 11.81 ( 0.32805 ) at  Asym = 42.981  xmid = 9.344  scal = 4.7286 f/j  13 / 7
## <<lamda: 0.065536  SS= 3.8548 ( 0.46972 ) at  Asym = 43.701  xmid = 6.3095  scal = 2.7564 f/j  15 / 8
## <<lamda: 0.026214  SS= 2.8386 ( 0.45514 ) at  Asym = 44.458  xmid = 5.503  scal = 2.0895 f/j  16 / 9
## <<lamda: 0.010486  SS= 2.4496 ( 0.2365 ) at  Asym = 46.866  xmid = 5.9479  scal = 2.3 f/j  17 / 10
## <<lamda: 0.0041943  SS= 2.0123 ( 0.21463 ) at  Asym = 51.378  xmid = 6.3516  scal = 2.3762 f/j  18 / 11
## <<lamda: 0.0016777  SS= 1.4462 ( 0.2703 ) at  Asym = 59.51  xmid = 7.03  scal = 2.5092 f/j  19 / 12
## <<lamda: 0.00067109  SS= 0.91272 ( 0.306 ) at  Asym = 71.787  xmid = 7.8952  scal = 2.6532 f/j  20 / 13
## <<lamda: 0.00026844  SS= 0.51816 ( 0.29993 ) at  Asym = 88.001  xmid = 8.8228  scal = 2.7857 f/j  21 / 14
## <<lamda: 0.00010737  SS= 0.26988 ( 0.25879 ) at  Asym = 107.49  xmid = 9.727  scal = 2.901 f/j  22 / 15
## <<lamda: 4.295e-05  SS= 0.13173 ( 0.20437 ) at  Asym = 129.22  xmid = 10.558  scal = 2.9978 f/j  23 / 16
## <<lamda: 1.718e-05  SS= 0.062773 ( 0.16799 ) at  Asym = 151.62  xmid = 11.279  scal = 3.0748 f/j  24 / 17
## <<lamda: 6.8719e-06  SS= 0.032464 ( 0.12324 ) at  Asym = 172.29  xmid = 11.849  scal = 3.1314 f/j  25 / 18
## <<lamda: 2.7488e-06  SS= 0.021923 ( 0.076781 ) at  Asym = 188.07  xmid = 12.235  scal = 3.1673 f/j  26 / 19
## <<lamda: 1.0995e-06  SS= 0.01977 ( 0.035429 ) at  Asym = 196.72  xmid = 12.43  scal = 3.1847 f/j  27 / 20
## <<lamda: 4.398e-07  SS= 0.019619 ( 0.0092109 ) at  Asym = 199.45  xmid = 12.49  scal = 3.19 f/j  28 / 21
## <<lamda: 1.7592e-07  SS= 0.019617 ( 0.00087595 ) at  Asym = 199.84  xmid = 12.499  scal = 3.1908 f/j  29 / 22
## <<lamda: 7.0369e-08  SS= 0.019617 ( 4.7993e-05 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  30 / 23
## <<lamda: 2.8147e-08  SS= 0.019617 ( 8.24e-07 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  31 / 24
\end{verbatim}

@@

The nonlinear modeling model specifications are, of course, a
development of linear ones. Unfortunately, the explicit model
\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the
linear modeling function \texttt{lm()}, which requires this model to be
specified as \texttt{y\ \textasciitilde{}\ x}.

Within \texttt{nls()}, consider the following FOUR different
specifications for the same problem, plus an intuitive choice, labelled
fm2a, that does not work. In this failed attempt, putting the
\texttt{Asym} parameter in the model causes the \texttt{plinear}
algorithm to try to add another term to the model. We believe this is
unfortunate, and would like to see a consistent syntax. At the time of
writing, we do not foresee a resolution for this issue. In the example,
we have NOT evaluated the commands to save space.

@@

@@

The \texttt{nls()} code is structured in a way that inhibits both
maintenance and improvement. In particular, the iterative setup is such
that introduction of Marquardt stabilization is not easily available. To
obtain performance when the code was written, a large portion was
written in C with consequent calls and returns that severely complicate
the code. Over time, R has become much more efficient on modern
computers, and the need to use compiled C and Fortran is less critical.
Moreover, the burden for maintenance could be reduced by moving code
entirely to R.

While R and its packages are generally very well documented for usage,
that documentation rarely extends as far as it might for code
maintenance. The paucity of such documentation is exacerbated by the
mixed R/C/Fortran code base, where seemingly trivial differences in
things like representation of numbers or base index for arrays lead to
traps for unwary programmers.

--\textgreater{} --\textgreater{} --\textgreater{} --\textgreater{}

@@

\begin{itemize}
\item
  to provide a packaged version of \texttt{nls()} (call it
  \texttt{nlsalt}) coded entirely in R that matches the features in base
  R or what is packaged in \texttt{nlspkg} as described in Nash and
  Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022}).
\item
  to streamline the overall \texttt{nls()} infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to
  maintenance of both the existing \texttt{nls()} methods and features
  as well as new features we or others would like to add.
\item
  to explain what we do, either in comments or separate maintainer
  documentation. Since we are complaining about the lack of explanatory
  material for the current code, we feel it is incumbent on us to
  provide such material for our own work, and if possible for the
  existing code.
\end{itemize}

@@

\begin{itemize}
\item
  A BibTex bibliography for use with all documents in this project, but
  which has wider application to nonlinear least squares projects in
  general
  (\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib})
\item
  \texttt{MachID.R} offers a suggested concise summary function to
  identify a particular computational system used for tests. A
  discussion of how it was built and the resources needed across
  platforms is given in at
  \url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}
\item
  Nash and Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022})
  is an explanation of the construction of the \texttt{nlspkg} from the
  \texttt{nls()} code in R-base.
\item
  As the 2021 Summer of Code period was ending, one of us (JN) was
  invited to prepare a review of optimization in R. Ideas from the
  present work have been instrumental in the creation of Nash
  (\protect\hyperlink{ref-NashWires22}{2022}).
\end{itemize}

@@

From a design point of view, a key difference in approach between
\texttt{nls()} and \texttt{nlsr::nlxb()} is that \texttt{nls()} builds a
large infrastructure, especially the \texttt{nlsModel()} function, from
which the Gauss-Newton iteration can be executed and other statistical
information such as profiles can be computed, while \texttt{nlxb()}
returns quite limited information, and computes what is needed on an
``as and when'' basis. This follows a path that one of us (JN)
established almost 50 years ago with the software that became Nash
(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
post-solution analysis phases of computations.

@@

\leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset
Orthogonality Convergence Criterion for Nonlinear Least Squares.''}
\emph{Technometrics} 23 (2): 179--83.

@@

\leavevmode\vadjust pre{\hypertarget{ref-nlsr-manual}{}}%
John C Nash, and Duncan Murdoch. 2019. \emph{{nlsr: Functions for
Nonlinear Least Squares Solutions}}.

@@
